Computational methods in landscape ecology

### 2024-09-14 Sections/03-connectivity.qmd
Draft V1

## Estimating connectivity
While connectivity is key to many modern conservation theories and practices, quantifying it has proven both technically and conceptually challenging (Calabrese & Fagan 2004; Taylor et al. 2006).  If we restrict ourselves to thinking about the dispersal of organisms from a patch in search of new breeding habitat (e.g., natal dispersal) then what do we need to measure to quantify connectivity? Ideally, we would have information for many organisms about the location they started from, high-resolution information about their movement trajectory and the location they ‘finished’ in (or their point of mortality if that occurred).  Depending on the question of interest, the endpoint of movement could represent a range of alternatives from the short-term movement to access food resources to finding and establishing a territory and successfully breeding. From this information, we may be able to infer the elements in the landscape that organisms prefer to move through and how their behaviour changes in them drawing on behavioural landscape ecology (Lima & Zollner 1996) and statistical tools such as statistical state-space models (Jonsen et al. 2013).  While high-resolution biotelemetry has made tracking animal movement at high-resolution possible (Kays et al. 2015), it is still logistically challenging to track many animals over long periods at high spatial resolution. Furthermore, we are often interested in predicting the consequences of changes in structural connectivity before they occur. Many disciplines (e.g., geography, archaeology, ecology, physics and chemistry) are interested in how matter and energy flow through heterogeneous media (including landscapes), and hence several methods have been developed to quantify connectivity.  As described above, because dispersal is a fundamental component of metapopulation, metacommunity and metaecosystem theory, adequately describing connectivity is vital for their empirical evaluation.

<!--
## Landscape metrics

There are several landscape metrics that are traditionally used to measure landscape connectivity.
For example, the habitat amount within a buffer zone of focal patches, the nearest-neighbor distance, or the connectance index [@Moilanen2002SimpleConnectivity; @Prugh2009EvaluationPatch; @Spanowicz2019MeasuringLandscape].
However, these metrics are frequently criticized for being to simple and especially the link to functional connectivity is often ambiguous. 
Thus, more functionality needs to be included using additional ecological meaningful metrics, such as the core area [e.g., @Kupfer2012LandscapeEcology].
More recently there is also an increasing focus in within-patch connectivity using metrics such as the effective mesh size in addition to the usual measure between-patch connectivity [@Spanowicz2019MeasuringLandscape].
-->

<!--
## Resistance surfaces

Resistance surfaces are one of the most popular approaches to model connectivity and developed from simple isolation-by-distance approaches. 
The approach uses environmental layers and biodiversity data or expert opinions to estimate the cost and of movement through landscapes [@Zeller2012EstimatingLandscape].
The latest developments of resistance surfaces have recently been reviewed [@Dutta2022OverviewComputational].
The resistance surface can be used to model movement based on different approaches such as leas-cost paths [@Adriaensen2003ApplicationLeastcost; @Etherington2016LeastcostModelling] or circuit theory [@McRae2008UsingCircuit].
However, resistance surfaces are not free of critique including missing spatial-temporal variability or context dependency [@UnnithanKumar2022ConnectivityModelling; @UnnithanKumar2022MovingLandscape].
Thus, recent developments include the combination with more processes-based modelling approaches [@Schumaker2018HexSimModeling; @Bocedi2021RangeShifterExtended; @UnnithanKumar2022PathwalkerNew], modelling of dynamic landscape connectivity [@Zeller2020UnderstandingImportance], inclusion of stochasticity and spatial context [@Ribeiro2017LandScapeCorridors], using genetic optimization algorithms [@Peterman2018ResistanceGAPackage], or computational improvements [@Leonard2017GlowSoftware].
Software to model connectivity based on resistance surfaces include *ConScape.jl* [@vanMoorter2023AcceleratingAdvances], *Omniscape.jl* [@Landau2021OmniscapeJl], *lconnect* [@Mestre2023LconnectPackage], *grainscape* [@Chubaty2020ToolboxGrainscape], or *Circuitscape* [@Anantharaman2019CircuitscapeJulia].
-->

<!-- What about multi-species approaches? -->

<!--
## Graph theory

Graph theory describes a landscape as a set of nodes representing habitat patches and edges as potential links between the nodes.
Thus, graph theory is a powerful tool to model landscape connectivity, especially in a metapopulation context [@Urban2001LandscapeConnectivity].

-->

<!--
Coupling landscape graph modeling and biological data https://doi.org/10.1007/s10980-020-00998-7
Temporal dynamics of landscape connectivity in ever-changing landscapes: a literature review https://doi.org/10.1007/s10980-021-01277-9
Some methodological considerations to improve landscape planning https://doi.org/10.1016/j.ecolind.2017.05.052
Comparison of landscape graph modelling methods https://doi.org/10.1007/s10980-020-01164-9

-->

<!--
## Other approaches
-->

<!--
A stochastic movement simulator improves estimates of landscape connectivity https://doi.org/10.1890/14-1690.1
Landscape connectivity modeling from the perspective of animal dispersal  https://doi.org/10.1007/s10980-019-00935-3
Big-data approaches https://doi.org/10.1126/science.abg1780
-->

>>>>

## Estimating connectivity
While connectivity is key to many modern conservation theories and practices, directly quantifying it has proven both technically and conceptually challenging [@Calabrese2004ConnectivityMetrics; @Taylor2006LandscapeConnectivity). Data from many individuals recording the start location, trajectory, and endpoint of a movement is required to directly observe connectivity [@Simpkins2018AssessingPerformance]. Using this data one may be able to infer the elements in the landscape that individuals preferentially move through. While constantly improving technlogies has made tracking animal movement at high-resolution possible [@Kays2015AnimalTracking], it is still logistically challenging to track high numbers of individuals for the required time periods and spatial resolution. As such in practice connectivity is primarily determined by indirect estimation. @Calabrese2004ConnectivityMetrics distinguised three types of connectivity estimates:
- Structural connectivity: determined by the physical attributes of the landscape (as described above).
- Potential connectivity: a combination of the physical attributes of the landscape and limited information about the dispersal capabilities of the target species, to estimate how connected a given landscape or patch is for that species.
- Actual connectivity: Related to observations of individuals moving through a landscape or between patches, providing solid estimates of the linkages between landscpae elements or habitat patches.

### Structural Connectivity Estimation Methods
Landscape metrics, produced using software such as `landscapemetrics` [@Hesselbarth2019LandscapemetricsOpen], are commonly used to quantify various aspects of landscape structure, decomposed into composition and configuration. These metrics describe structural connectivity in terms of:
- Amount of habitat: Total area of habitat types present.
- Patch size: Average size of habitat patches.
- Patch isolation: Average nearest-neighbour distance between patches.
- Connectance index: the percentage of all possible connections between habitat patches that actually exist in a landscape given a pre-defined maximum distance between patches.

More recently there has been an increasing focus on within-patch connectivity using metrics such as the effective mesh size in addition to the usual metrics [@Spanowicz2019MeasuringLandscape].

### Estimating Functional Connectivity Using Resistance Surfaces
Functional connectivity estimation involves integrating species-specific movement data with landscape features. This estimation normally involves the usage of graph-theoretic approaches, though these details are often obscured for convenience by software [@McRae2008UsingCircuit]. Two primary base approaches are usually applied in modern connectivity estimation:
- Least-cost paths (LCP)
- Circuit theory

A large number of software applications have been developed, with these approaches as a base, to estimate connectivity including *ConScape.jl* [@vanMoorter2023AcceleratingAdvances], *Omniscape.jl* [@Landau2021OmniscapeJl], *lconnect* [@Mestre2023LconnectPackage], *grainscape* [@Chubaty2020ToolboxGrainscape], or *Circuitscape* [@Anantharaman2019CircuitscapeJulia].

These methods all rely on resistance surfaces to represent the landscape [@Zeller2012EstimatingLandscape]. A resistance surface is a grid-based representation of a landscape where each cell is assigned a value representing the difficulty (or cost) for an individual of a particular species to traverse that cell [@Etherington2016LeastcostModelling]. The values are based on landscape features and represent factors influencing movement, such as habitat type, topography, or barriers.

Creating a resistance surface involves:
- Collecting data: Obtaining landscape data for the area of interest along with a sufficiently large buffer area to avoid edge artifacts [@Dutta2022OverviewComputational].
- Categorisation: Grouping landscape grids into thematically similar categories to capture key elements impacting species movements.
- Assigning resistance or cost values: Quantifying the value for each cell in the grid using empirical movement data, which can include:
  - Expert opinion: Very widely used due to its ubiquity and low cost, however the accuracy of using this approach is difficult to measure resulting in uncertainties [@Krueger2012ExpertOpinion].
  - Detection data: Single point locations of unknown individuals.
  - Relocation data: Multiple sequential locations of the same individual but at low frequency (e.g. mark recapture).
  - Pathway data: High-frequency relocation data allowing for movement track inference (e.g., GPS tracking).
  - Genetic data: Genetic samples used to calculate genetic distances between populations.

Resistance surfaces are not free of critique including missing spatial-temporal variability or context dependency [@UnnithanKumar2022ConnectivityModelling; @UnnithanKumar2022MovingLandscape]. Thus, recent developments include the combination with more processes-based modelling approaches [@Schumaker2018HexSimModeling; @Bocedi2021RangeShifterExtended; @UnnithanKumar2022PathwalkerNew], modelling of dynamic landscape connectivity [@Zeller2020UnderstandingImportance], inclusion of stochasticity and spatial context [@Ribeiro2017LandScapeCorridors], using genetic optimization algorithms [@Peterman2018ResistanceGAPackage], or computational improvements [@Leonard2017GlowSoftware].

#### Least-cost paths (LCP)
LCP is among the most widely used methods for quantifying functional connectivity. The underlying assumption is that organisms will move through landscapes in ways that minimize the cumulative cost of traversing different landscape elements. LCP models assume that individuals have perfect knowledge of the landscape and will choose the path that minimizes resistance costs perfectly with no deviations [@Adriaensen2003LCP]. The creation of LCPs is a well established technique and as such there are highly optimised algorithms available for its calculation, often making the approach more computationally effient than others

By calculating the LCP between two points using a resistance surface, researchers can estimate connectivity between these points based on the accumulated cost along the path. This method can also generate accumulated cost surfaces depicting the minimum cost from a single point to all other locations, helping identify reachable areas within a cost threshold.

#### Circuit theory
Circuit theory considers all possible pathways between locations simultaneously [@McRae2008UsingCircuit]. In this model, the landscape is represented as a network where nodes are connected by resistors (weighted by the values in the resistance surface), and movement is analogous to electrical current flow. Circuit theory does not assume that organisms have perfect knowledge of the landscape and is particularly useful when multiple alternative pathways are available.

Key measures from circuit theory are:
- Resistance distance: The isolation between pairs of locations measured by incorporating multiple pathways. Locations connected by many low-resistance pathways have lower resistance distances.
- Current flow: Identifies areas of high movement probability, highlighting important corridors and bottlenecks in the landscape.
